# Multilevel linear models: the basics (Ch 12)

This chapter starts on page 251

```{r, include=FALSE, warning=F, message=F}
# knit options
knitr::opts_chunk$set(
  echo = TRUE
  , warning = FALSE
  , message = FALSE
  , results='hide'
  , fig.width = 10
  , fig.height = 7
)
# qualitative color pallette
my_color_pal_qual <- function(n) viridis::turbo(n = n * 4)[which(seq(1,n * 4,1) %% 4 == 0)]
```

## Setup

```{r, eval=T}
# bread-and-butter
library(tidyverse)
library(lubridate)
library(viridis)
library(scales)
library(latex2exp)
# visualization
library(cowplot)
library(kableExtra)
# Linear Mixed-Effects Models 
library(lme4)
library(broom.mixed)
# jags and bayesian
library(rjags)
library(MCMCvis)
library(HDInterval)
library(BayesNSF)
#set seed
set.seed(11)
```
 
## Load data

```{r, results='asis'}
# srrs2 <- read.table ("../data/radon/srrs2.dat", header=TRUE, sep=",")
srrs2 <- read.table ("http://www.stat.columbia.edu/~gelman/arm/examples/radon/srrs2.dat", header=TRUE, sep=",")
srrs2 %>% 
  dplyr::glimpse()
# filter MN and create vars
radon_mn <- srrs2 %>% 
  dplyr::filter(
    state=="MN"
  ) %>% 
  dplyr::mutate(
    radon = activity
    , log.radon = log(ifelse(radon==0, .1, radon))
    , y = log.radon
    , x = floor # 0 for basement, 1 for first floor
    , county_index = as.numeric(as.factor(county))
  )
# n
n <- nrow(radon_mn)
# count
radon_mn %>% dplyr::count(county_index, county) %>% dplyr::slice_head(n=10) %>% 
  kableExtra::kable(
    caption = "Records by county (first 10)"
  ) %>% 
  kableExtra::kable_styling(font_size = 12) %>% 
  kableExtra::column_spec(1, bold = FALSE, width = "3em")
```

## Varying-intercept multilevel model, no predictors (p. 253)

$$
y_i = \alpha_{j[i]} + \epsilon_i
$$

where $j[i]$ is the county corresponding to house $i$

Partial-pooling estimates from a multilevel model: 
For this simple scenario with no predictors, the multilevel estimate for a given county $j$ can be approximated as a weighted average of the mean of the observations in the county (the unpooled estimate, $\overline{y}_j$) and the mean over all counties (the completely pooled estimate, $\overline{y}_{all}$).

The weighted average reflects the relative amount of information available about the individual county, on one hand, and the average of all the counties, on the other: 

* Averages from counties with smaller sample sizes carry less information, and the weighting pulls the multilevel estimates closer to the overall state average. In the limit, if $n_j = 0$, the multilevel estimate is simply the overall average, $\overline{y}_{all}$.
* Averages from counties with larger sample sizes carry more information, and the corresponding multilevel estimates are close to the county averages. In the limit as $n_j \to \infty$, the multilevel estimate is simply the county average, $\overline{y}_{j}$ . 
* In intermediate cases, the multilevel estimate lies between the two extremes.

### Data and Initial conditions

```{r}
# varying-intercept model, no predictors
## data for jags
data <- list(
  n = nrow(radon_mn)
  , J = radon_mn$county %>% unique() %>% length()
  , y = radon_mn$log.radon %>% as.double()
  , county = radon_mn$county_index %>% as.double()
)
## inits for jags
inits <- function(){
  list(
    alpha = rnorm(n = data$J, mean = 0, sd = 1) # county-level intercept (n = length of # counties)
    , mu.alpha = rnorm(n = 1, mean = 0, sd = 1)
    , sigma.alpha = runif(n = 1, min = 0, max = 1)
    , sigma.y = runif(n = 1, min = 0, max = 1)
  )
}
# define parameters to return from MCMC
params <- c ("alpha", "mu.alpha", "sigma.y", "sigma.alpha")
```

### JAGS Model

Write out the JAGS code for the model.

```{r, eval=FALSE}
## JAGS Model
model {
  ###########################
  # priors
  ###########################
  # alpha priors
    mu.alpha ~ dnorm(0, .0001)
    tau.alpha <- 1/sigma.alpha^2
    sigma.alpha ~ dunif(0, 100)
    for (j in 1:J){
      alpha[j] ~ dnorm(mu.alpha, tau.alpha)
    }
  # y priors
    sigma.y ~ dunif (0, 100)
    tau.y <- 1/sigma.y^2
  ###########################
  # likelihood
  ###########################
  for (i in 1:n){
    y[i] ~ dnorm(alpha[county[i]], tau.y)
  }
}
```

### Implement JAGS Model

```{r}
##################################################################
# insert JAGS model code into an R script
##################################################################
{ # Extra bracket needed only for R markdown files - see answers
  sink("intrcpt_nopred.R") # This is the file name for the jags code
  cat("
  ## JAGS Model
  model {
    ###########################
    # priors
    ###########################
    # alpha priors
      mu.alpha ~ dnorm(0, .0001)
      tau.alpha <- 1/sigma.alpha^2
      sigma.alpha ~ dunif(0, 100)
      for (j in 1:J){
        alpha[j] ~ dnorm(mu.alpha, tau.alpha)
      }
    # y priors
      sigma.y ~ dunif (0, 100)
      tau.y <- 1/sigma.y^2
    ###########################
    # likelihood
    ###########################
    for (i in 1:n){
      y[i] ~ dnorm(alpha[county[i]], tau.y)
    }
  }
  ", fill = TRUE)
  sink()
}
##################################################################
# implement model
##################################################################
######################
# Call to JAGS
######################
intrcpt_nopred = rjags::jags.model(
  file = "intrcpt_nopred.R"
  , data = data
  , inits = inits
  , n.chains = 3
  , n.adapt = 100
)
stats::update(intrcpt_nopred, n.iter = 1000, progress.bar = "none")
# save the coda object (more precisely, an mcmc.list object) to R
mlm_intrcpt_nopred = rjags::coda.samples(
  model = intrcpt_nopred
  , variable.names = params
  , n.iter = 1000
  , n.thin = 1
  , progress.bar = "none"
)
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

### Summary of the marginal posterior distributions of the parameters

```{r, results='asis'}
# summary
MCMCvis::MCMCsummary(mlm_intrcpt_nopred, params = params[params != "alpha"]) %>% 
  kableExtra::kable(
    caption = "Bayesian: arying-intercept multilevel model, no predictors"
    , digits = 5
  ) %>% 
  kableExtra::kable_styling(font_size = 12)
```

### Make a horizontal caterpillar plot for the $\alpha_{j}$

```{r}
# Caterpillar plots
MCMCvis::MCMCplot(
  mlm_intrcpt_nopred
  , params = c("alpha")
  , horiz = FALSE
  , ylim = c(-6,5)
  # Number specifying size of text for parameter labels on axis.
  , sz_labels = 0.6
  # Number specifying size of points represents posterior medians.
  , sz_med = 0.7
  # Number specifying thickness of 50 percent CI line (thicker line).
  , sz_thick = 2
  # Number specifying thickness of 95 percent CI line (thinner line).
  , sz_thin = 1
)
```

### Sort the $\alpha_{j}$ by number of obs.

Replicate Figure 12.1 (b) (p.253)

```{r}
dta_temp <- dplyr::bind_cols(
  radon_mn %>% dplyr::count(county_index) %>% data.frame()
  , MCMCvis::MCMCpstr(mlm_intrcpt_nopred, params = "alpha", func = function(x) quantile(x, c(0.25, 0.5, 0.75))) %>%
      data.frame() %>% 
      dplyr::rename_with(
        ~ tolower(gsub(".", "", .x, fixed = TRUE))
      )
  ) %>% 
  dplyr::arrange(n) %>% 
  dplyr::mutate(n_low_hi = dplyr::row_number() %>% as.factor())
ggplot(dta_temp) +
  geom_hline(
    yintercept = MCMCvis::MCMCpstr(mlm_intrcpt_nopred, params = "mu.alpha", func = median) %>% unlist()
    , linetype = "dashed"
    , lwd = 1
    , color = "gray40"
  ) +
  geom_point(
    mapping = aes(x = n_low_hi, y = alpha50)
  ) +
  geom_linerange(
    mapping = aes(x = n_low_hi, ymin = alpha25, ymax = alpha75)
  ) +
  scale_y_continuous(limits = c(0,2.5)) +
  ylab(latex2exp::TeX("$\\hat{\\alpha}_{j}$")) + 
  xlab("# Obs.") +
  labs(
    title = "Median model intercept prediction for each county (j)"
    , subtitle = "Varying-intercept multilevel model, no predictors"
  ) +
  theme_bw() + 
  theme(
    axis.text.x = element_blank()
    , axis.ticks.x = element_blank()
    , panel.grid = element_blank()
  )
```


## Classical no-pooling with predictors (p. 254)

This section considers partial pooling for a model with unit-level predictors. In this scenario, no pooling might refer to fitting a separate regression model within each group. However, a less extreme and more common option that we also sometimes refer to as "no pooling" is a model that includes group indicators and estimates the model classically.

### Fit no-pooling model with `lm`

$$
y_i = \alpha_{j[i]} + \beta\cdot x_i + \epsilon_i
$$

where $j[i]$ is the county corresponding to house $i$; and $x$ is the floor of measurement (0 = basement, 1 = 1st floor...)

To fit the no-pooling model in R, we include the county index (a variable named county that takes on values between 1 and 85) as a factor in the regression—thus, predictors for the 85 different counties. We add “−1” to the regression formula to remove the constant term, so that all 85 counties are included. Otherwise, R would use county 1 as a baseline. 

In the no-pooling model, the $\alpha_{j}$'s are set to the classical least squares estimates, which correspond to the fitted intercepts in a model run separately in each county (with the constraint that the slope coefficient equals $\beta$ in all models).

```{r, results='asis'}
lm(data = radon_mn, formula = y ~ x + factor(county_index) - 1) %>% 
  broom::tidy() %>% 
  dplyr::slice_head(n=10) %>% 
  kableExtra::kable(
    caption = "no-pooling model includes county indicators"
  ) %>% 
  kableExtra::kable_styling(font_size = 12) %>% 
  kableExtra::column_spec(1, bold = FALSE, width = "3em")
```

## MLM Partial pooling with predictors (p. 254)

The simplest multilevel model for the radon data with the floor predictor can be written as:

\begin{align*}
y_i \sim \textrm{N}(\alpha_{j[i]} + \beta\cdot x_i, \sigma^2) \\
\textrm{for} &\; i = 1, \cdots, n
\end{align*}

which looks like the no-pooling model but with one key difference. In the no-pooling model, the $\alpha_{j}$'s are set to the classical least squares estimates, which correspond to the fitted intercepts in a model run separately in each county (with the constraint that the slope coefficient equals $\beta$ in all models). In the multilevel model, a "soft constraint" is applied to the $\alpha_{j}$'s: they are assigned a probability distribution,

\begin{align*}
\alpha_j &\sim \textrm{N}(\mu_{\alpha}, \sigma^{2}_{\alpha}) \\
\textrm{for} &\; j = 1, \cdots, J
\end{align*}

with their mean $\mu_{\alpha}$ and standard deviation $\sigma^{2}_{\alpha}$ estimated from the data. The distribution (12.3) has the effect of pulling the estimates of $\alpha_{j}$ toward the mean level $\mu_{\alpha}$, but not all the way. (p. 257)

### Using `lme4::lmer`

```{r, results='asis'}
## Varying-intercept model w/ no predictors
M0 <- lme4::lmer(data = radon_mn, formula = y ~ 1 + (1 | county))
summary(M0)

## Including x as a predictor
M1 <- lme4::lmer(data = radon_mn, formula = y ~ x + (1 | county))
summary(M1)

#estimated regression coefficicents
coef(M1)$county[1:3,1]

# fixed and random effects
fixef(M1)
ranef(M1)$county[1:3,1]

# check
fixef(M1)[[1]] + ranef(M1)$county[1,1]
coef(M1)$county[1,1]
```

### Bayesian

#### Data and Initial conditions

```{r}
# varying-intercept model, no predictors
## data for jags
data <- list(
  n = nrow(radon_mn)
  , J = radon_mn$county %>% unique() %>% length()
  , y = radon_mn$log.radon %>% as.double()
  , x = radon_mn$x %>% as.double()
  , county = radon_mn$county_index %>% as.double()
)
## inits for jags
inits <- function(){
  list(
    alpha = rnorm(n = data$J, mean = 0, sd = 1) # county-level intercept (n = length of # counties)
    , beta = rnorm(1, mean = 0, sd = 1)
    , mu.alpha = rnorm(n = 1, mean = 0, sd = 1)
    , sigma.alpha = runif(n = 1, min = 0, max = 1)
    , sigma.y = runif(n = 1, min = 0, max = 1)
  )
}
# define parameters to return from MCMC
params <- c ("alpha", "beta", "mu.alpha", "sigma.y", "sigma.alpha")
```

#### JAGS Model

Write out the JAGS code for the model.

```{r, eval=FALSE}
## JAGS Model
model {
  ###########################
  # priors
  ###########################
  # alpha priors
    mu.alpha ~ dnorm(0, .0001)
    tau.alpha <- 1/sigma.alpha^2
    sigma.alpha ~ dunif(0, 100)
    for (j in 1:J){
      alpha[j] ~ dnorm(mu.alpha, tau.alpha)
    }
  # beta
    beta ~ dnorm(0, .0001)
  # y priors
    sigma.y ~ dunif (0, 100)
    tau.y <- 1/sigma.y^2
  ###########################
  # likelihood
  ###########################
  for (i in 1:n){
    mu.y[i] <- alpha[county[i]] + beta * x[i]
    y[i] ~ dnorm(mu.y[i], tau.y)
  }
}
```

#### Implement JAGS Model

```{r}
##################################################################
# insert JAGS model code into an R script
##################################################################
{ # Extra bracket needed only for R markdown files - see answers
  sink("intrcpt_pred.R") # This is the file name for the jags code
  cat("
  ## JAGS Model
  model {
    ###########################
    # priors
    ###########################
    # alpha priors
      mu.alpha ~ dnorm(0, .0001)
      tau.alpha <- 1/sigma.alpha^2
      sigma.alpha ~ dunif(0, 100)
      for (j in 1:J){
        alpha[j] ~ dnorm(mu.alpha, tau.alpha)
      }
    # beta
      beta ~ dnorm(0, .0001)
    # y priors
      sigma.y ~ dunif (0, 100)
      tau.y <- 1/sigma.y^2
    ###########################
    # likelihood
    ###########################
    for (i in 1:n){
      mu.y[i] <- alpha[county[i]] + beta * x[i]
      y[i] ~ dnorm(mu.y[i], tau.y)
    }
  }
  ", fill = TRUE)
  sink()
}
##################################################################
# implement model
##################################################################
######################
# Call to JAGS
######################
intrcpt_pred = rjags::jags.model(
  file = "intrcpt_pred.R"
  , data = data
  , inits = inits
  , n.chains = 3
  , n.adapt = 100
)
stats::update(intrcpt_pred, n.iter = 1000, progress.bar = "none")
# save the coda object (more precisely, an mcmc.list object) to R
mlm_intrcpt_pred = rjags::coda.samples(
  model = intrcpt_pred
  , variable.names = params
  , n.iter = 1000
  , n.thin = 1
  , progress.bar = "none"
)
```

```{r, warning=FALSE, message=FALSE, echo=FALSE, include=FALSE}
remove(list = ls()[grep("_temp",ls())])
gc()
```

#### Summary of the marginal posterior distributions of the parameters

```{r, results='asis'}
# summary
MCMCvis::MCMCsummary(mlm_intrcpt_pred, params = params[params != "alpha"]) %>% 
  kableExtra::kable(
    caption = "Bayesian: Varying-intercept multilevel model, floor of measurement predictor"
    , digits = 5
  ) %>% 
  kableExtra::kable_styling(font_size = 12)
```

Compare fully Bayesian approach to linear mixed-effects model (via Restricted Maximum Likelihood [REML]) 

```{r, results='asis'}
# compare
MCMCvis::MCMCsummary(mlm_intrcpt_pred, params = "alpha") %>% 
  data.frame() %>% 
  dplyr::select("mean") %>% 
  dplyr::slice_head(n = 8) %>% 
  dplyr::bind_cols(coef(M1)$county[1:8,1]) %>% 
  dplyr::rename(Bayesian=1, LMM=2) %>% 
  kableExtra::kable(
    caption = "alpha predictions Bayesian vs. LMM"
    , digits = 5
  ) %>% 
  kableExtra::kable_styling(font_size = 12)
```

#### Sort the $\alpha_{j}$ by number of obs.

Replicate Figure 12.1 (b) (p.253)

```{r}
dta_temp <- dplyr::bind_cols(
  radon_mn %>% dplyr::count(county_index) %>% data.frame()
  , MCMCvis::MCMCpstr(mlm_intrcpt_pred, params = "alpha", func = function(x) quantile(x, c(0.25, 0.5, 0.75))) %>%
      data.frame() %>% 
      dplyr::rename_with(
        ~ tolower(gsub(".", "", .x, fixed = TRUE))
      )
  ) %>% 
  dplyr::arrange(n) %>% 
  dplyr::mutate(n_low_hi = dplyr::row_number() %>% as.factor())
ggplot(dta_temp) +
  geom_hline(
    yintercept = MCMCvis::MCMCpstr(mlm_intrcpt_pred, params = "mu.alpha", func = median) %>% unlist()
    , linetype = "dashed"
    , lwd = 1
    , color = "gray40"
  ) +
  geom_point(
    mapping = aes(x = n_low_hi, y = alpha50)
  ) +
  geom_linerange(
    mapping = aes(x = n_low_hi, ymin = alpha25, ymax = alpha75)
  ) +
  scale_y_continuous(limits = c(0,2.5)) +
  ylab(latex2exp::TeX("$\\hat{\\alpha}_{j}$")) + 
  xlab("# Obs.") +
  labs(
    title = "Median model intercept prediction for each county (j)"
    , subtitle = "Varying-intercept multilevel model, floor of measurement predictor"
  ) +
  theme_bw() + 
  theme(
    axis.text.x = element_blank()
    , axis.ticks.x = element_blank()
    , panel.grid = element_blank()
  )
```
