# Multilevel linear models: varying slopes, non-nested models, and other complexities (Ch 13)

This chapter starts on page 279

```{r, include=FALSE, warning=F, message=F}
# knit options
knitr::opts_chunk$set(
  echo = TRUE
  , warning = FALSE
  , message = FALSE
  # , results='hide'
  , fig.width = 10
  , fig.height = 7
)
```

## Setup

```{r, eval=T}
# bread-and-butter
library(tidyverse)
library(lubridate)
library(viridis)
library(scales)
library(latex2exp)
# visualization
library(cowplot)
library(kableExtra)
# Linear Mixed-Effects Models 
library(lme4)
library(broom.mixed)
# jags and bayesian
library(rjags)
library(MCMCvis)
library(HDInterval)
library(BayesNSF)
#set seed
set.seed(11)
```
 
## Load data

```{r, results='asis'}
# load individual, house-level observations
srrs2 <- read.table ("http://www.stat.columbia.edu/~gelman/arm/examples/radon/srrs2.dat", header=TRUE, sep=",") %>% 
  dplyr::mutate(
    fullfips = stfips*1000 + cntyfips
  )
# load county data
cty <- read.table ("http://www.stat.columbia.edu/~gelman/arm/examples/radon/cty.dat", header=TRUE, sep=",") %>% 
  dplyr::mutate(fullfips = stfips*1000 + ctfips) %>% 
  # there are duplicates in this data
  dplyr::group_by(fullfips) %>% 
  dplyr::filter(dplyr::row_number()==1) %>% 
  dplyr::ungroup() %>% 
  dplyr::mutate(u = log(Uppm))
# join to individual, house radon measurement data
srrs2 <- srrs2 %>% 
  dplyr::left_join(cty %>% dplyr::select(fullfips, u), by = c("fullfips"="fullfips"))
# filter MN and create vars
radon_mn <- srrs2 %>% 
  dplyr::filter(
    state=="MN"
  ) %>% 
  dplyr::mutate(
    radon = activity
    , log.radon = log(ifelse(radon==0, .1, radon))
    , y = log.radon
    , x = floor # 0 for basement, 1 for first floor
    , county_index = as.numeric(as.factor(county))
  )
# unique uranium by county index 1:85
u <- radon_mn %>% 
  dplyr::group_by(county_index) %>% 
  summarize(u = dplyr::first(u)) %>% 
  ungroup() %>% 
  dplyr::arrange(county_index)
```

## Varying intercepts and slopes (p. 279)

The next step in multilevel modeling is to allow more than one regression coefficient to vary by group. We shall illustrate with the radon model from the previous chapter, which is relatively simple because it only has a single individual-level predictor, $x$ (the indicator for whether the measurement was taken on the first floor). We begin with a varying-intercept, varying-slope model including $x$ but without the county-level uranium predictor; thus, 

\begin{align*}
y_i &\sim \textrm{N}(\alpha_{j[i]} + \beta_{j[i]} \cdot x_i, \sigma^2_y) \\
&\textrm{for} \; i = 1, \cdots, n \\
\left(\begin{array}{c}
\alpha_{j}\\
\beta_{j}
\end{array}\right)
&\sim \textrm{N}\left(\left(\begin{array}{c}
\mu_\alpha\\
\mu_\beta
\end{array}\right) ,
\left(\begin{array}{cc}
\sigma^{2}_{\alpha} & \rho\sigma_{\alpha}\sigma_{\beta}\\
\rho\sigma_{\alpha}\sigma_{\beta} & \sigma^{2}_{\beta}
\end{array}\right)
\right) \\
&\textrm{for} \; j = 1, \cdots, J
\end{align*}

with variation in the $\alpha_{j}$’s and the $\beta_{j}$’s and also a between-group correlation parameter $\rho$

### Using `lme4::lmer` (p. 279)

```{r, results='asis'}
## Varying intercept & slopes w/ no group level predictors
M3 <- lmer (data = radon_mn, y ~ x + (1 + x | county))
summary(M3)

#estimated regression coefficicents
coef(M3)$county[1:3,]

# fixed and random effects
fixef(M3)
ranef(M3)$county[1:3,]

# check intercept match for county 1
fixef(M3)[[1]] + ranef(M3)$county[1,1]
coef(M3)$county[1,1]
```

